---
title: "5100 Final Project T-Test"
author: "Dana Kim"
format: 
  html:
    embed-resources: true
toc: true
---

# CLEANING THE DATA

## T-test after removing null values for sex demographic

Reading in the data with only relevant, renamed columns and deleting all columns that have null values for any of the three demographic categories (race, sex, age).

```{r}
library(tidyverse)
complaints_nonull_demographic <- read_csv("../../../data/cleaning/null_demographics_removed.csv")
complaints_nonull_demographic
```

After removing all null values in any of the three demographic categories (age, sex, race), we individually removed all unknown values from our individual demographic categories. For this analysis, I removed the "U" values from the sex category (assuming that "U" stands for unknown).

```{r}
#CLEAN TO REMOVE UNKNOWN SEX VALUES
#remove "U" values in age column assuming they mean unknown - confirm that this is how Jackson did null bootstrap analysis*

sex_nonull_nounknown <- complaints_nonull_demographic |>
  filter(susp_sex != "U")

sex_nonull_nounknown
nrow(sex_nonull_nounknown) #315,344



#VISUALIZE (for all precincts before vs. after)
ggplot(data=sex_nonull_nounknown, aes(x=time_period, fill=susp_sex)) +
  geom_bar()
```

We then find the top 20 precincts with the most complaints after removing the nulls from any demographic category (race, sex, gender). This keeps the top 20 precincts among sex, race, and gender the same. These are the precincts we will run our analyses on.

```{r}
#FIND TOP 20 PRECINCTS WITH THE MOST COMPLAINTS FROM NO NULL DEMOGRAPHIC

top_20_precincts <- complaints_nonull_demographic |>
  count(precinct_crime, sort = TRUE) |>  
  slice_max(n, n = 20)              

top_20_precincts$precinct_crime


#filter to keep only top 20 precincts after removing null from 3 demographics
sex_nonull_nounknown_top20 <- sex_nonull_nounknown |>
  filter(precinct_crime %in% top_20_precincts$precinct_crime)
sex_nonull_nounknown_top20

table(sex_nonull_nounknown_top20$precinct_crime)

#VISUALIZE (for top 20 precincts before vs. after)
ggplot(data=sex_nonull_nounknown_top20, aes(x=time_period, fill=susp_sex)) +
  geom_bar()



#CREATE COPY TO RENAME DATAFRAME FOR CLARITY
sex_df <- sex_nonull_nounknown_top20
nrow(sex_df) #143,930
```

Now we run the t-test. First, a little bit more exploratory data analysis to see what the sample sizes are.

```{r}
#T-TEST
#for each precinct, is the difference in suspect sex before vs. after patternizr significant?
library(tidyverse)
library(lubridate)

#SEPARATE BY TIME PERIOD
#separate data based on before and after patternizr
before <- sex_df |>
  filter(time_period == "before Patternizr")
head(before)

after <- sex_df |>
  filter(time_period == "after Patternizr")
head(after)

#BEFORE PATTERNIZR
#for each precinct, what is the average monthly number of male and female suspects before patternizr?
(table(before$precinct_crime, before$susp_sex))

#AFTER PATTERNIZR
#for each precinct, what is the average monthly number of male and female suspects after patternizr?
(table(after$precinct_crime, after$susp_sex))

#table confirms that for t-test assumption of normality, we can proceed since the sample sizes are large enough to withstand any deviations in normality. don't need to check for variance sine welch's t-test is done below
```


# Z-test

## Z-test to see whether PROPORTION of that sex committing crime relative to other sexes in each time period is statistically different before vs after Patternizr

Where the proportion is given by a vector of 1s and 0s. where 1 is true and 0 is false indicating whether in the before/after period that sex committed a crime. Mean of those 1/0s are equal to the percentage of sex (ie. males) who commmitted crime before vs after

```{r}
### DEFINE FUNCTION TO ANALYZE ONE PRECINCT'S MALE VS. FEMALE INCREASE AFTER PATTERNIZER

ztest_precinct <- function(precinct_df) {
  #SET UP
  sex_levels <- unique(precinct_df$susp_sex) #get unique sexes, just M and F
  sex_results <- list() #create empty list to store results from each sex's t-test
  
  precinct_name <- unique(precinct_df$precinct_crime)[1] #get precinct name to label results accordingly
  
  #separate precinct's df by time period
  before <- precinct_df |> filter(time_period == "before Patternizr")
  after <- precinct_df |> filter(time_period == "after Patternizr")
  
  #LOOP
  for (current_sex in sex_levels){
  #add column to each before/after df to tell whether sex of interest (in for loop iteration) is found
  before$committed <- as.numeric(before$susp_sex == current_sex)
  after$committed <- as.numeric(after$susp_sex == current_sex)
  #print(before)
  #print(after)
  
  #t-test for whether percent of sex of interest (ie male) committed crimes is greater after than before
  zt <- prop.test(
      x = c(sum(after$committed), sum(before$committed)), # Count of Successes
      n = c(length(after$committed), length(before$committed)), # Count of Total Trials
      alternative = "greater",
      correct = FALSE # FALSE makes it a standard Z-test
    )
  
  #save result of sex (ie male) t-test to list
  sex_results[[current_sex]] <- data.frame(
    precinct       = precinct_name,
    sex            = current_sex,
    prop_before    = zt$estimate[2], 
    prop_after     = zt$estimate[1],
    p_value        = zt$p.value,
    n_before       = length(before$committed), # Total crimes in 'before' period
    n_after        = length(after$committed)   # Total crimes in 'after' period
  )
  }

#COMBINE RESULTS FOR THIS PRECINCT
precinct_results <- bind_rows(sex_results)
return(precinct_results)
}

### APPLY THE FUNCTION TO ALL PRECINCTS
precinct_list <- split(sex_df, sex_df$precinct_crime) #subset the sex_df into separate precincts

all_precincts_list <- lapply(precinct_list, ztest_precinct) #apply to all precincts listed
all_precincts_df <- bind_rows(all_precincts_list)
print(all_precincts_df)

### EXPORT RESULTS TO CSV
write_csv(all_precincts_df, "sex_z_test.csv")
```
```{r}
#significant results
significant_precincts <- all_precincts_df |>
  filter(p_value < 0.05)
significant_precincts
```

```{r}
#Testing the z-test function
#test_df <- bind_rows(lapply(precinct_list[1], ztest_precinct))
#print(test_df)
```
# Z-test results

INSERT INTERPRETATION OF RESULTS HERE.

Interestingly, from these results we can see that only the proportion of females to males increased at a statistically significant rate after the use of Patternizr for all but one precinct. This is likely due to the fact that before Patternizr, females were already not in the complaint records as often as males. However, since we know that the overall monthly average complaints for males and females increased after Patternizr, the results from this test suggest that this increase was statistically significant even in relation to males. However, males are still overwhelmingly complained about at a higher rate than females. Given this, it makes sense why only the female to male ratio was found to be statistically significant but still points to an important finding that after the use of Patternizr the proportion of females to males increased but the portion of males to females did not increase.