---
title: "EDA, T-Test, and Chi-Square Tests - Age"
author: "Fareeza Ali"
format:
  html:
    embed-resources: true
    fig-width: 10
    fig-height: 6
toc: true
---

Conducted on different precincts than previously.


```{r}
library(tidyverse)
library(ggplot2)
library(plotly)

#read in and drop rows with unknown values in age column
df <- read.csv("../../data/cleaning/null_demographics_removed.csv", stringsAsFactors = FALSE) |> 
  filter(susp_age_group != "UNKNOWN") |>
  mutate(complaint_date = as.Date(complaint_date))


```

We filtered to the 20 precincts with the most crime data reported in our cleaned dataset (including before and after Patternizr).

Note: 77,272 UNKNOWN age values were removed \[discuss effect on the dataset\]

```{r}
#rank top20 precincts
precinct_rank <- df |> count(precinct_crime, name = "n") |>
    arrange(desc(n)) |>
    mutate(rank = dense_rank(desc(n))
    ) 

top_precincts <- precinct_rank |> 
    slice(1:20) |>
    pull(precinct_crime
    )

top_precincts
```

```{r}
# EDA: table and scatterplot
age_table <- df |>
  count(susp_age_group, precinct_crime, time_period, name = "n") |>
  arrange(desc(n)
  )

age_table 

age_wide <- age_table |>
  filter(precinct_crime %in% top_precincts) |>
  tidyr::pivot_wider(
    names_from  = time_period,
    values_from = n,
    values_fill = 0
  ) |>
  rename(
    before = `before Patternizr`,
    after  = `after Patternizr`
  )

p <- ggplot(age_wide,
       aes(x = before, y = after, color = susp_age_group,
           text = paste(
             "Precinct:", precinct_crime,
             "<br>Age Group:", susp_age_group,
             "<br>Before:", scales::comma(before),
             "<br>After:", scales::comma(after)
           ))) +
  geom_point(alpha = 0.6, size = 3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(
    title = "Arrest Counts by Age Group (All Precincts): Before vs After Patternizr",
    x = "Before Patternizr (arrest count)",
    y = "After Patternizr (arrest count)",
    color = "Age Group"
  ) +
  theme_minimal(base_size = 14)

# Convert to interactive plotly widget
ggplotly(p, tooltip = "text")


```

The above table and scatterplot shows the breakout of each age group in each of the top 20 precincts, both before and after Patternizr. Dots above the line show more arrests in that age group for that precinct after Patternizr. Dots below the line show fewer arrests. The further from the line, the larger the shift was.

We can see that almost all age groups had ore arrests after Patternizr than before. For age group/precinct combos with less arrests after Patternizr, the decrease is small compares to the increases. The largest overall increases seem to be in the 24-44 range, and the decreases seem to be concentrated within the \<18 and 18-24 age groups.

```{r}
# EDA: Heatmap
df |> filter(precinct_crime %in% top_precincts) |>
  count(susp_age_group, precinct_crime, time_period) |>
  ggplot(aes(precinct_crime, susp_age_group, fill = n)) +
  geom_tile() +
  scale_fill_viridis_c(option = "C", trans = "log10", breaks = c(1,10,100,1000)) +
  facet_wrap(vars(time_period)) +
  labs(x = "Precinct crime", y = "Suspect age group", fill = "Count") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
# EDA: stackplots
df |>
  filter(precinct_crime %in% top_precincts) |>
  count(time_period, precinct_crime, susp_age_group) |>
  ggplot(aes(precinct_crime, n, fill = susp_age_group)) +
  geom_col(position = "stack") +
  facet_wrap(vars(time_period), scales = "free_y") +
  labs(x = "Precinct crime", y = "Count", fill = "Suspect age") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
# EDA: Histograms
df |>
  filter(precinct_crime %in% top_precincts) |>
  mutate(precinct_crime = factor(precinct_crime, levels = top_precincts)) |>
  count(precinct_crime, susp_age_group, time_period) |>
  ggplot(aes(x = susp_age_group, y = n, fill = time_period)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  facet_wrap(vars(precinct_crime), scales = "free_y", drop = FALSE) +
  labs(x = "Suspect age group", y = "Count", fill = "Time period") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

This above plots display arrest counts by age group in the top precincts, comparing the periods before and after Patternizr. Each panel represents one precinct, and the bars show the number of arrests for each suspect age group, broken down by time period.

Across nearly all precincts, the patterns of bar heights show large and systematic changes in the age distribution of arrests after Patternizr was introduced. Three major shifts are visible: - Arrests of younger suspects (\<18 and 18–24) generally decreased after Patternizr, suggesting Patternizr is associated with fewer arrests of minors and young adults across many top precincts. - Arrests of older suspects (25–44 and especially 45–64) increased after Patternizr, suggesting Patternizr is associated with substantial increases in arrests of adults aged 25–64. - The shift is not isolated and appears across nearly all top precincts, suggesting the effect is not specific to a precint but systematic, meaning Patternizr changes how arrests are distributed demographically, not just the volume of arrests in a few locations.

## T-test

```{r}
library(lubridate)
#t-tests for each precinct based on monthly averages

# t tests for each precinct
df_tp <- df |>
  filter(precinct_crime %in% top_precincts) |>
  mutate(time_period = factor(time_period,
                              levels = c("before Patternizr", "after Patternizr")))

# get age groups and precincts
precincts   <- df_tp |> distinct(precinct_crime) |> pull()
age_levels  <- df_tp |> distinct(susp_age_group) |> pull()

# create empty tiblble for results
age_ttests_by_precinct <- tibble()

# nested loop -> precint then age
for (p in precincts) {
  for (age in age_levels) {

    #filter to precinct + age group
    df_p <- df_tp |> filter(precinct_crime == p, susp_age_group == age)

    #create monthly data + count monthly arrests
    monthly_counts <- df_p |> 
      mutate(month = floor_date(complaint_date, "month")) |>
      count(time_period, month)

    # skip for groups with 0 counts 
    if (monthly_counts |> count(time_period) |> nrow() < 2) next

    #skip if either period has zero variance (constant monthly counts)
    group_vars <- monthly_counts |> 
      group_by(time_period) |> 
      summarize(v = var(n), .groups = "drop")
    if (any(is.na(group_vars$v)) || any(group_vars$v == 0)) next

    #monthly count t-tests
    tt <- t.test(n ~ time_period, data = monthly_counts)

    #append results
    age_ttests_by_precinct <- age_ttests_by_precinct |>
      bind_rows(tibble(
        precinct_crime = p,
        susp_age_group = age,
        mean_before    = tt$estimate[1],
        mean_after     = tt$estimate[2],
        diff_mean      = tt$estimate[2] - tt$estimate[1],
        statistic      = tt$statistic,
        p_value        = tt$p.value,
        conf_low       = tt$conf.int[1],
        conf_high      = tt$conf.int[2],
        n_months       = nrow(monthly_counts)
      ))
  }
}


#check results
age_ttests_by_precinct

#write.csv(age_ttests_by_precinct, "age_precinct_t-tests_final.csv", row.names = FALSE)

```

## Z-test 

H₀: The proportion of arrests in a given age group is the same before and after Patternizr. H₁: The proportion of arrests in a given age group is less than or greater after Patternizr than before.

```{r}
#precincts with statistically significant differences in age group arrests
sig_results <- age_ttests_by_precinct |>
  filter(p_value < 0.05)

sig_results

#unique precincts in dataset
sig_precincts <- sig_results|> distinct(precinct_crime) |> pull()
sig_precincts

#write.csv(sig_results, "age_precinct_t-tests_significant.csv", row.names = FALSE)
```

Acrosss the statistically significant age group - precinct combinations (where p-value < 0.05), the introduction of Patternizr was associated with a substantial increase in monthly arrests. We see the strongest effects on the 25-44 age group, with individual precinct increases from ~ 15-18 additional arrests each month. Arrests in the 18-24 age group also increased significantly by ~ 5-7 additional arrests per month. The 45-64 age group had a moderate increase of ~4-5 additional arrests per month. Overall, the significant results show a consistent demographic shift after Patternizr: arrests of adults increased substantially across the top precincts.

Precincts 40 and 75 indicate the strongest overall effect, with each having more than 40 additional arrests per month across age groups after the introduction of Patternizr. Precincts 41, 44, and 47 also had substantial increases. These precincts' effects are mostly driven by a rise in arrests of the 25-44 age group.

```{r}
# z tests for each precinct
df_tp <- df |>
  filter(precinct_crime %in% top_precincts) |>
  mutate(time_period = factor(time_period,
                              levels = c("before Patternizr", "after Patternizr")))

# get age groups and precincts
precincts   <- df_tp |> distinct(precinct_crime) |> pull()
age_levels  <- df_tp |> distinct(susp_age_group) |> pull()

# create empty tiblble for results
age_ztests_by_precinct <- tibble()

# nested loop -> precint then age
for (p in precincts) {
  df_p <- df_tp |> filter(precinct_crime == p)
  for (age in age_levels) {
    tmp <- df_p |>
      mutate(ind = susp_age_group == age)

    # counts
    x_before <- sum(tmp$ind[tmp$time_period == "before Patternizr"])
    x_after  <- sum(tmp$ind[tmp$time_period == "after Patternizr"])

    n_before <- sum(tmp$time_period == "before Patternizr")
    n_after  <- sum(tmp$time_period == "after Patternizr")

    # proportions
    p_before <- x_before / n_before
    p_after  <- x_after  / n_after

    # pooled proportion
    p_pool <- (x_before + x_after) / (n_before + n_after)

    # standard error
    se <- sqrt(p_pool * (1 - p_pool) * (1/n_before + 1/n_after))

    # z statistic
    z_stat <- (p_after - p_before) / se

    # two-sided p-value
    p_val <- 2 * (1 - pnorm(abs(z_stat)))
    
    # append results
    age_ztests_by_precinct <- age_ztests_by_precinct |>
      bind_rows(tibble(
        precinct_crime = p,
        susp_age_group = age,
        estimate_diff  = p_after - p_before,
        statistic      = z_stat,
        p_value        = p_val,
        n_before       = n_before,
        n_after        = n_after
      ))

  }
}

#check results
age_ztests_by_precinct

write.csv(age_ztests_by_precinct, "age_precinct_z-tests_final.csv", row.names = FALSE)
```

```{r}
#precincts with statistically significant differences in age group arrests
sig_results_z <- age_ztests_by_precinct |>
  filter(p_value < 0.05)

sig_results_z

#unique precincts in dataset
sig_precincts_z <- sig_results_z|> distinct(precinct_crime) |> pull()
sig_precincts_z

#write.csv(sig_results_z, "age_precinct_z-tests_significant.csv", row.names = FALSE)
```
## Chi Square test

```{r}
#chi square test - is the distribution of one categorical variable different across levels of another categorical variable?
#is the count of age group level different depending on the time period?

#test 1: did age distribution change time_period ~ sus_age_group
df_tp_cs <- df |>
  filter(precinct_crime %in% top_precincts)

precincts[1] 
head(df_tp_cs)

chi_results <- tibble()

for (p in precincts) {
  # Subset to this precinct
  df_p <- df_tp_cs |> filter(precinct_crime == p)

  # Create contingency table: age group by before/after
  tab <- table(df_p$susp_age_group, df_p$time_period)
  print(tab)
  
  # Run chi-square test
  chi <- chisq.test(tab)
  print(chi)
  
  # Store results
  chi_results <- chi_results |>
    bind_rows(tibble(
      precinct_crime = p,
      statistic = as.numeric(chi$statistic),
      p_value = chi$p.value,
      df = chi$parameter
    ))
}

chi_results
#write.csv(chi_results, "age_precinct_chi_results.csv", row.names = FALSE)

```

```{r}
#precincts with statistically significant differences chi results
sig_chi_results <- chi_results |>
  filter(p_value < 0.05)

sig_chi_results

#unique precincts in dataset
sig_chi_results_precincts  <- sig_chi_results |> distinct(precinct_crime) |> pull()
sig_chi_results_precincts

#write.csv(sig_chi_results_precincts, "age_precinct_chi_result_significant.csv", row.names = FALSE)
```
# RESUME


```{r}
# Make sure df_tp is already filtered to top precincts and before/after
df_tp_cs <- df |>
  filter(precinct_crime %in% top_precincts)

posthoc_results <- tibble()

for (p in precincts[1]) {
  
  df_p <- df_tp_cs |> filter(precinct_crime == p)
  print(df_p)
  
  # Build age × time table
  tab <- table(df_p$susp_age_group, df_p$time_period)
  print(tab)
  
  # Chi-square test
  chi <- chisq.test(tab)
  print(chi)
  
  # Standardized residuals matrix
  resid_mat <- chi$stdres
  
  # Convert matrix → tidy tibble
  resid_df <- as.data.frame(as.table(resid_mat)) |>
    as_tibble() |>
    rename(
      susp_age_group = Var1,
      time_period    = Var2,
      std_residual   = Freq
    ) |>
    mutate(
      precinct_crime = p,
      chi_statistic  = as.numeric(chi$statistic),
      p_value        = chi$p.value
    )
  
  # Append
  posthoc_results <- posthoc_results |> bind_rows(resid_df)
}

posthoc_results

#write.csv(posthoc_results, "chi_posthoc_results.csv", row.names = FALSE)
```




```{r}
# Create wide-format before/after table
shift_results <- posthoc_results |>
  select(precinct_crime, susp_age_group, time_period, std_residual) |>
  mutate(time_period = ifelse(time_period == "before Patternizr", "before", "after")) |>
  tidyr::pivot_wider(
    names_from = time_period,
    values_from = std_residual
  ) |>
  mutate(
    shift = after - before   # positive = more overrepresented after
  )
 
shift_results |> filter(shift > 2)

shift_results |> filter(shift < -2)

shift_results |> arrange(desc(abs(shift))) |> head(20)


#visualize
library(ggplot2)

ggplot(shift_results,
       aes(x = "Before", y = before,
           xend = "After", yend = after,
           group = susp_age_group,
           color = susp_age_group)) +
  geom_segment(size = 1.1, alpha = 0.8) +
  geom_point(size = 3, aes(x = "Before", y = before)) +
  geom_point(size = 3, aes(x = "After",  y = after)) +
  facet_wrap(~ precinct_crime, scales = "free_y") +
  labs(
    title = "Standardized Residual Shifts: Before vs After Patternizr",
    x = "",
    y = "Standardized Residual",
    color = "Age Group"
  ) +
  theme_minimal() +
  theme(
    panel.spacing = unit(1, "lines"),
    axis.text.x = element_text(face = "bold")
  )
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```