
```{r}
# ============================================
# Race-based monthly t-tests by precinct
# Before vs After Patternizr
# ============================================

library(dplyr)
library(tidyr)
library(tibble)
library(lubridate)

# --------------------------------------------
# 1. Load and clean race data
# --------------------------------------------

data <- read.csv("../../../data/cleaning/null_demographics_removed.csv")

# Remove UNKNOWN race
data <- data[data$susp_race != "UNKNOWN", ]


# Convert complaint_date to Date
# If your dates look like "04/15/2022", use mdy()
# If they look like "2022-04-15", use ymd() instead.
#data$complaint_date <- mdy(data$complaint_date)
data$complaint_date <- ymd(data$complaint_date)  # <-- use this instead if needed
data
```

```{r}
# --------------------------------------------
# 2. Get the top 20 precincts
# --------------------------------------------

top_precincts_table <- data %>%
  count(precinct_crime, sort = TRUE) %>%
  slice_max(n, n = 20) %>%
  arrange(precinct_crime)

print(top_precincts_table)

top_precincts <- top_precincts_table$precinct_crime

# --------------------------------------------
# 3. Filter to top precincts & prep for t-tests
# --------------------------------------------

df_tp_race <- data %>%
  filter(precinct_crime %in% top_precincts) %>%
  mutate(
    time_period = factor(
      time_period,
      levels = c("before Patternizr", "after Patternizr")
    )
  )
df_tp_race

# Get list of precincts and race categories
precincts   <- df_tp_race %>% distinct(precinct_crime) %>% pull()
precincts

race_levels <- df_tp_race %>% distinct(susp_race) %>% pull()
race_levels
```

```{r}
#check the assumptions of normality and equal variance

# --------------------------------------------
# 3b. Check Assumptions (Normality & Variance)
# --------------------------------------------

assumption_results <- tibble()

for (p in precincts) {
  for (race in race_levels) {
    
    # 1. PREPARE DATA (Same aggregation as your T-test loop)
    df_p <- df_tp_race %>%
      filter(precinct_crime == p,
             susp_race == race)
    
    if (nrow(df_p) == 0) next
    
    # Create monthly counts
    monthly_counts <- df_p %>%
      mutate(month = floor_date(complaint_date, "month")) %>%
      count(time_period, month) %>%
      ungroup()
    #print(monthly_counts)
    
    # Split into vectors for testing
    counts_before <- monthly_counts %>% 
      filter(time_period == "before Patternizr") %>% pull(n)
    
    counts_after <- monthly_counts %>% 
      filter(time_period == "after Patternizr") %>% pull(n)
    
    # 2. RUN DIAGNOSTICS
    # Initialize variables as NA
    shapiro_before_p <- NA
    shapiro_after_p  <- NA
    var_test_p       <- NA
    
    # Check Normality (Requires n >= 3 and valid variance)
    # If p < 0.05, Normality is VIOLATED
    if (length(counts_before) >= 3 && var(counts_before) > 0) {
      shapiro_before_p <- shapiro.test(counts_before)$p.value
    }
    
    if (length(counts_after) >= 3 && var(counts_after) > 0) {
      shapiro_after_p <- shapiro.test(counts_after)$p.value
    }
    
    # Check Equal Variance (Requires data in both groups)
    # If p < 0.05, Equal Variance is VIOLATED
    if (length(counts_before) >= 2 && length(counts_after) >= 2 && 
        var(counts_before) > 0 && var(counts_after) > 0) {
      var_test_p <- var.test(counts_before, counts_after)$p.value
    }
    
    # 3. STORE RESULTS
    assumption_results <- assumption_results %>%
      bind_rows(tibble(
        precinct_crime = p,
        susp_race      = race,
        n_months_pre   = length(counts_before),
        n_months_post  = length(counts_after),
        normality_pre_p  = shapiro_before_p,
        normality_post_p = shapiro_after_p,
        equal_var_p      = var_test_p
      ))
  }
}

# --------------------------------------------
# VIEW VIOLATIONS
# --------------------------------------------

# Identify groups where Normality is violated (p < 0.05)
non_normal_groups <- assumption_results %>%
  filter(normality_pre_p < 0.05 | normality_post_p < 0.05)

# Identify groups where Equal Variance is violated (p < 0.05)
unequal_variance_groups <- assumption_results %>%
  filter(equal_var_p < 0.05)

print(paste("Total tests run:", nrow(assumption_results)))
print(paste("Groups violating normality:", nrow(non_normal_groups)))
print(paste("Groups violating equal variance:", nrow(unequal_variance_groups)))

# View the table
assumption_results
```
Although normality was violated, we have enough "months" of data per precinct per race for MOST precinct/race specific datasets (ie monthly count of a given precinct, for a given race). However, there are a handful that do not have enough "months" of data to overcome this. For these specific races in these specific precincts, we should conduct a non-parametric alternative like the Wilcoxon rank test. For the sake of time, we will continue with the Welch's t-test which accounts for unequal variance.

For many of the precinct and race specific datasets, we see that the equal variance assumptions was violated. We conducted a welch's t-test by default for all precincts based on these results.

```{r}
# --------------------------------------------
# 4. Monthly t-tests for each precinct × race
# --------------------------------------------

race_ttests_by_precinct <- tibble()

for (p in precincts) {
  for (race in race_levels) {
    
    # Filter to precinct + race
    df_p <- df_tp_race %>%
      filter(precinct_crime == p,
             susp_race == race)
    
    # If no rows, skip
    #if (nrow(df_p) == 0) next
    
    # Build monthly counts of arrests
    monthly_counts <- df_p %>%
      mutate(month = floor_date(complaint_date, "month")) %>%
      count(time_period, month)
    #print(monthly_counts)
    
    if (nrow(monthly_counts) < 30) next
    
    # Need data in both time periods
    #if (monthly_counts %>% count(time_period) %>% nrow() < 2) next
    
    # Need non-zero variance in both groups to run t-test
    group_vars <- monthly_counts %>%
      group_by(time_period) %>%
      summarize(v = var(n), .groups = "drop")
    
    if (any(is.na(group_vars$v)) || any(group_vars$v == 0)) next
    
    # Run t-test: monthly counts before vs after
    tt <- t.test(n ~ time_period, data = monthly_counts, var.equal=FALSE)
    
    # Make sure we map means correctly to "before" and "after"
    ests <- tt$estimate
    mean_before <- as.numeric(ests["before Patternizr"])
    mean_after  <- as.numeric(ests["after Patternizr"])
    
    # Append results
    race_ttests_by_precinct <- race_ttests_by_precinct %>%
      bind_rows(tibble(
        precinct_crime = p,
        susp_race      = race,
        mean_before    = mean_before,
        mean_after     = mean_after,
        diff_mean      = mean_after - mean_before,
        statistic      = as.numeric(tt$statistic),
        p_value        = tt$p.value,
        conf_low       = tt$conf.int[1],
        conf_high      = tt$conf.int[2],
        n_months       = nrow(monthly_counts)
      ))
  }
}
```


```{r}
# View full results
race_ttests_by_precinct

# Save full results (all tests)
#write.csv(
#  race_ttests_by_precinct,
#  "/Users/ishika/Desktop/MS/DSAN 5100/Final Project/T-tests/race_precinct_t-tests.csv",
#  row.names = FALSE
#)

# 5. Significant results (p < 0.05)

sig_race_ttests <- race_ttests_by_precinct %>%
  filter(p_value < 0.05)

sig_race_ttests

sig_precincts_race_t <- sig_race_ttests %>%
  distinct(precinct_crime) %>%
  pull()

sig_precincts_race_t

#write.csv(
#  sig_race_ttests,
#  "/Users/ishika/Desktop/MS/DSAN 5100/Final Project/T-tests/race_precinct_t-tests_significant.csv",
#  row.names = FALSE
#)
```

```{r}
significant_count <- sig_race_ttests |>
  count(susp_race)
significant_count

library(ggplot2)

# 2. Create the bar plot with fill by race
sig_count_plot <- ggplot(significant_count, aes(x = susp_race, y = n, fill = susp_race)) + # <--- Changed here
  geom_bar(stat = "identity") +
  
  # Add the count labels
  geom_text(aes(label = n), vjust = -0.5, color = "black") +
  
  # Fix cut-off at the top
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  
  # Titles and Labels
  labs(title = "Count of Significant Change in Monthly Complaints after Patternizr",
       x = "Suspect Race",
       y = "Count",
       fill = "Race") + # Legend title
  
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
sig_count_plot

ggsave("ttest_frequency.png", plot = sig_count_plot, width = 10, height = 6, units = "in", dpi = 300)
```

